---
title: "Import"
author: "Trevor S"
---

## Setup

We install and load the necessary packages.

```{r, warnings = FALSE, message = FALSE}
# install.packages("rvest")
# install.packages("stringr")
# install.packages("tidyverse")
# install.packages("janitor")
library(rvest)
library(stringr)
library(tidyverse)
library(janitor)
```

## Introduction

In the import step we import the data required for this report. As mentioned before, we will be importing data from [Hockey Reference](https://www.hockey-reference.com/draft). Before we import any data, it's important to consider *which* and *how many* years we want to include in this analysis. Since the NHL has changed dramatically over the years, care must be taken to ensure we do not include drafts from too long ago. The primary concern with including data from too many years ago is that teams have likely changed their drafting approach over time. For example, teams may have become better at evaluating prospects as more advanced statistics have been developed, meaning that there are likely fewer late round draft "steals" in the 2020s than there were in the 1980s. Thus including drafts from the 1980s would skew our calculations because it would overestimate contributions by players who were drafted in the later rounds, since those players would potentially have been drafted sooner if the teams of the 1980s had the resources available to teams today. This would make our model a poor estimator of draft pick value for drafts occurring in the 2020s. That being said, players drafted in recent years have not had sufficient time to contribute to their teams, so we should not include drafts from too recently either. Ideally, we would wait until all players from a draft class have retired before including it in our analysis . Practically speaking, this is not feasible since players can have very long careers (for example, Alex Ovechkin was drafted in 2004 and is still playing) which would force us to include older drafts to maintain the same sample size, which is also not ideal as explained above.

Having considered this, we make the somewhat arbitrary decision to use the 25 drafts between and 1996 and 2020 (inclusive). Note that a significant portion of the players in our dataset are still active, so we will have to make an adjustment to account for this. Additionally, it makes sense to give more recent drafts more weight for the reasons described above. We will make both of these adjustments in the [Transform chapter](https://trevsteu.github.io/Stat468_Final/transform.html).

## Code

We start off by creating a function to import data from Hockey Reference.

```{r, message=FALSE, warnings = FALSE, warning = FALSE}
start_year <- 1996
end_year <- 2020

import_draft <- function(year){
  url <- str_glue("https://www.hockey-reference.com/draft/NHL_{year}_entry.html")
  html <- read_html(url)
  Sys.sleep(5) # to avoid getting rate limited
  draft_year_table <- html |> 
    html_element("table") |> 
    html_table() |> 
    janitor::row_to_names(1) |> 
    janitor::clean_names()
  draft_year_table
}

head(import_draft(start_year), 10)
```

We compare the first 10 rows of the 1996 draft table shown above with the table on [Hockey Reference](https://www.hockey-reference.com/draft/NHL_1996_entry.html) It seems that the function we created does what we want it to do. We can now proceed to the Tidy step.
