---
title: "Tidy"
author: "Trevor S"
---

```{r, echo=FALSE, hide=TRUE, message=FALSE, warning=FALSE}
library(rvest)
library(tidyverse)
library(janitor)
start_year <- 1996
end_year <- 2020

import_draft <- function(year){
  url <- str_glue("https://www.hockey-reference.com/draft/NHL_{year}_entry.html")
  html <- read_html(url)
  draft_year_table <- html |> 
    html_element("table") |> 
    html_table() |> 
    janitor::row_to_names(1) |> 
    janitor::clean_names()
  draft_year_table
}
```

## Introduction

In the tidy step, we put the data into tidy form and clean it, which will make the data easier to analyze in the later steps. Despite the table from the previous chapter *looking* fairly clean, further inspection reveals some issues:

```{r, warning = FALSE}
import_draft(start_year)[23:30,]
```

Three problems that immediately come up are:

-   Two rows get inserted at the end of every round to indicate the round changed.

-   Numbers are being coded as strings (`overall`, `age`, `to`, etc).

-   (At least) one player is missing everything except for their pick number, name, team, position, nationality, and amateur team.

By doing a little bit of detective work with some of the other players with missing values elsewhere in the dataset, we notice that players who never played in the NHL have empty strings listed for everything except for the values attributes listed above. We will have to deal with this in the tidy step. Note that Hockey Reference begins listing player's ages in the 2001 draft, but we aren't going to use ages for our analysis so we won't bother coming up for a remedy for the players drafted between 1996 and 2000. Finally, it would be helpful to remove the columns we don't care about.

## Setup

```{r}
# install.packages("tidyverse")
library(tidyverse)
```

## Code

We build a function to tidy the data. In particular, we want it to:

-   Remove the rows added between rounds.

-   Correct the types of each column so we can use numeric columns in calculations.

-   Change `gp` and `ps` values to 0 for players who never played in the NHL or have a negative `ps`.

-   If `is.na(to)`, then the player never played in the NHL, so set it to the draft year.

-   Add a `year` column so we can adjust the stats of players drafted more recently.

-   Select the columns we care about (`year`, `overall`, `pos`, `to`, `gp`, and `ps`) in that order.

Note that there were originally two `gp` columns (one for games played and one for games played as a goalie, goalies have the same number in both), but when we used `janitor::clean_names()` it changed them to `gp` and `gp_2`. Additionally, we cannot remove the round separating rows by removing a specified row number since many of the drafts in our dataset have different numbers of picks per round, and some rounds within the same draft have even had a different numbers of picks.

```{r, warning = FALSE, message = FALSE}
tidy_draft <- function(year){
  draft_year_table <- import_draft(year) |> 
    filter(overall != "Overall" & overall != "")|> # remove extra rows
    type_convert() |> 
    mutate("year" = year, "ps" = pmax(coalesce(ps, 0), 0), 
           "gp" = coalesce(gp, 0), "to" = coalesce(to, year)) |> 
    select(year, overall, to, pos, gp, ps)
  draft_year_table
}

tidy_draft(1996)
```

This is the form we will use for analysis later. We now load in all of the data and use `rbind()` to bind the tables together, giving us a single data frame to work with.
