---
title: "Visualize"
author: "Trevor S"
---

## Setup

We install and load the necessary packages, along with functions from prior chapters.

```{r,  message = FALSE, warning = FALSE}
# renv::install("gt")
# renv::install("ggplot2")
# renv::install("patchwork")

library(gt)
library(ggplot2)
library(patchwork)

source("functions.R") # load functions defined in prior chapters
```

## Introduction

In the visualize step, we will perform some EDA (Exploratory Data Analysis) to get a sense of what our data looks like. Specifically, we will see if there are any patterns or trends that may be useful in the Model chapter.

## Exploratory Data Analysis (EDA)

### Number of Picks

Recall that over the years the NHL has changed how many picks there are in each round as franchises have been added. A consequence of this is that the number of rounds has also changed, and the number of total picks in a draft has changed several times throughout our dataset. Recall that we removed all picks after #224, but there could be drafts with *fewer* than 224 total selections. We check for this:

```{r}
all_data |> 
  group_by(year) |> 
  summarize(num_picks = n()) |> 
  ggplot(aes(x = year, y = num_picks)) + # takes less space than gt()
  geom_point()
```

Indeed the drafts after 2005 all have fewer than 224 selections. This is not a major problem since those picks aren't worth very much anyway, but it is worth noting that several picks from #211 onward have a smaller sample size than picks 1-210.

### PS and GP Values

Before doing any further EDA, we will take the five number summary, mean, and standard deviation of both the GP and PS values to get a sense of what they look like. Recall that the five number summary gives the minimum, 25% quantile, median, 75% quantile, and maximum of a dataset. Additionally, recall that PS stands for Point Share, and is a measure of a player's career contributions to points in the standings (ie the points you get from wins, not the one that is goals plus assists).

```{r}
c("five num" = fivenum(all_data$gp), "mean" = mean(all_data$gp), "sd" = sd(all_data$gp))

c("five_num" = fivenum(all_data$ps), "mean" = mean(all_data$ps), "sd" = sd(all_data$ps))
```

Clearly both the GP and PS values are right skewed. Note that the maximum of the GP data is around 6 standard deviations from the mean $(\frac{1779-142.7713}{274.0303} = 5.97)$ , whereas the maximum of the PS data is almost 10 standard deviations away $(\frac{217.8-8.249862}{21.172591} = 9.89)$.

We next check what proportion of our dataset ever played in an NHL game and what proportion generated less than 2 PS in their career, which is the value of exactly one win in the NHL.

```{r}
all_data |> 
  filter(gp > 0) |> 
  nrow() / nrow(all_data)

all_data |> 
  filter(ps < 2) |> 
  nrow() / nrow(all_data)
```

This tells us that just over half of our dataset never played in an NHL game and almost three quarters made minimal on-ice contributions in their career.

To get visual confirmation that our data is very right skewed, we check the histograms of the data, one of GP (on the left) and one of PS (on the right). We also set the scales to be the same to make comparing the values easier.

```{r, message=FALSE, warning = FALSE}
gp_hist <- all_data |> 
  ggplot(aes(gp)) + 
  geom_histogram() + 
  scale_y_continuous(limits = c(0, 4200)) + 
  labs(title = "Distribution of GP", 
       x = "GP", y = "Number of Players") 

ps_hist <- all_data |> 
  ggplot(aes(ps)) + 
  geom_histogram() + 
  scale_y_continuous(limits = c(0, 4200)) + 
  labs(title = "Distribution of PS", 
       x = "PS", y = "Number of Players") 

gp_hist + ps_hist
```

Indeed, both of these are very right-skewed, and clearly a lot of players end up playing a small number of games and are thus not generating able to generate much PS. The fact that there are more players with a small PS than a small GP also makes sense since players could be unproductive in \~75 games, which would take them out of the first bin for GP while they remain ib the first bin for PS.

### Picking a Metric

We may also guess that GP and PS are positively correlated, since better players get to play in more games and thus accumulate more PS. Indeed:

```{r}
cor(all_data$ps, all_data$gp)
```

Because of this, we choose to only include one of GP and PS in each model to avoid multicollinearity concerns. As mentioned in the Approach section, we will use two GP-related metrics and two PS-related metrics. For each of these metrics we will fit a linear and non-linear regression model, for a total of 8 models. We now explore the GP and PS values in greater detail.

### GP and PS of Draft Classes

We wish to confirm that players selected earlier in a draft (ie a lower `overall`) tend to play in more games and generate more PS during their careers than those selected later. To check this, we start by creating plot of the GP and PS values by overall for a single draft. Note we will use `scales = "free"` because the scales of GP and PS are quite different:

```{r}
set.seed(468) # for reproducibility
rand_year <- sample(start_year:end_year, 1) # year is 2015

gp_plot_1 <- all_data |> 
  filter(year == rand_year) |> 
  ggplot(aes(x = overall, y = gp)) +
  geom_point() +
  labs(x = "Pick Number", y = "GP", 
       title = str_glue("GP of Players Drafted in {rand_year}"), 
       subtitle = "Note the plots have different scales")

ps_plot_1 <- all_data |> 
  filter(year == rand_year) |> 
  ggplot(aes(x = overall, y = ps)) +
  geom_point() +
  labs(x = "Pick Number", y = "PS", 
       title = str_glue("PS of Players Drafted in {rand_year}"), 
       subtitle = "Note the plots have different scales")

gp_plot_1 + ps_plot_1
```

Since clearly so many players play in 0 games and thus generate 0 PS, we will recreate the plot without the players that played in 0 NHL games to make the plot easier to read. This time we will include multiple drafts.

```{r}
gp_plot_2 <- all_data |> 
  filter(gp > 0) |> 
  ggplot(aes(x = overall, y = gp)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Distribution of GP",
       subtitle = "Players with ≥ 1 game only; \nnote the plots have different scales", 
       x = "GP", y = "Number of Players")

ps_plot_2 <- all_data |> 
  filter(gp > 0) |> 
  ggplot(aes(x = overall, y = ps)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Distribution of PS",
       subtitle = "Players with ≥ 1 game only; \nnote the plots have different scales", 
       x = "PS", y = "Number of Players")

gp_plot_2 + ps_plot_2
```

These plots are quite dense and difficult to interpret, but there isn't really any point in jittering the data because it'll still overlap, so we re plot them with a random 5 year sample of our dataset.

```{r}
years <- sample(start_year:end_year, 5) # years are 2018, 1996, 1999, 2010, 2004

gp_plot_3 <- all_data |> 
  filter(gp > 0 & year %in% years) |> 
  ggplot(aes(x = overall, y = gp)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Distribution of GP for 5 Drafts",
       subtitle = "Players with ≥ 1 game only; \nnote the plots have different scales", 
       x = "GP", y = "Number of Players")

ps_plot_3 <- all_data |> 
  filter(gp > 0 & year %in% years) |> 
  ggplot(aes(x = overall, y = ps)) + 
  geom_point(alpha = 0.5) + 
  labs(title = "Distribution of PS for 5 Drafts",
       subtitle = "Players with ≥ 1 game only; \nnote the plots have different scales", 
       x = "PS", y = "Number of Players")

gp_plot_3 + ps_plot_3 
```

Though these plot are still quite busy, it shows a strange trend that there seems to be more players drafted around 200 overall that end up having successful careers, than drafted around 125. I am not sure of an underlying reason for this, but we will need to be careful when modeling to ensure that these late picks are not given more value than earlier picks.

The plots above are not great because they are missing the vast majority of our dataset (20 drafts plus all the players who played in 0 NHL games for the 5 years in the sample). To improve this, we plot the mean GP and mean PS of the players selected at each pick. Given how skewed our data is, typically the median would be preferred over the mean to improve resistance to outliers. However in this context we don't really want to be resistant to outliers since the outliers represent elite players (those who had a lot of GP or PS), and ignoring those players would make our model underestimate the values of picks which have been used to select extremely successful players.

```{r, warning=FALSE}
gp_plot_4 <- all_data |> 
  group_by(overall) |> 
  summarize(mean_gp = mean(gp)) |> 
  ggplot(aes(x = overall, y = mean_gp)) + 
  geom_point() +
  geom_point(aes(x = 156, y = mean(filter(all_data, overall==156)$gp)), col = "dodgerblue") +
  geom_point(aes(x = 205, y = mean(filter(all_data, overall==205)$gp)), col = "salmon") +
  labs(title = "Mean GP by Pick Number", subtitle = "Note the scales are different",
       x = "Pick Number", y = "Mean GP") + 
  annotate(geom = "segment", x = 156, y = 450, xend = 156, yend = 220, colour = "dodgerblue",
    arrow = arrow(type = "open", length = unit(0.32, "cm"))) +
  annotate(geom = "label", x = 90, y = 400,
    label = "156th overall selection,\n(mean GP of 203)",
    hjust = "left", colour = "dodgerblue")

ps_plot_4 <- all_data |> 
  group_by(overall) |> 
  summarize(mean_ps = mean(ps)) |> 
  ggplot(aes(x = overall, y = mean_ps)) + 
  geom_point() +
  geom_point(aes(x = 156, y = mean(filter(all_data, overall==156)$ps)), col = "dodgerblue") +
  geom_point(aes(x = 205, y = mean(filter(all_data, overall==205)$ps)), col = "salmon") +
  labs(title = "Mean PS by Pick Number", subtitle = "Note the scales are different",
       x = "Pick Number", y = "Mean PS") + 
  annotate(geom = "segment", x = 175, y = 37.5, xend = 203, yend = 14, colour = "salmon",
    arrow = arrow(type = "open", length = unit(0.32, "cm"))) +
  annotate(geom = "label", x = 100, y = 39,
    label = "205th overall selection,\n(mean PS of 13.124)",
    hjust = "left", colour = "salmon")

gp_plot_4 + ps_plot_4
```

Interestingly, points that appear to be outliers in mean GP may not be outliers in mean PS (and vise versa). This is indicated by the blue and pink points in each plot. We can also see that in general GP and PS both tend to decrease later in drafts, and that pick value tends to level off around pick 75, though there are some picks that stick out (for example, pick 205 has an average PS of 13.124, whereas pick 204 has an average PS of 4.856). This particular outlier is due to Henrik Lundqvist and Joe Pavelski being selected at this spot and having career PSs of 173.3 and 130.1, respectively. Very few players drafted this late make it to the NHL (20 of the 25 players in our dataset drafted at pick 205 have more than 0.3 PS), so two players with very successful careers skewing the GP and PS data is not surprising.

## Summary

We take the following lessons from our EDA into our Transform and Model chapters:

-   We will only use one of GP or PS in each model to avoid multicollinearity-related issues.

-   We will need to fit a curve or line to the data, since if we simply set $v_i$ to be the mean GP or PS of players selected at pick $i$ then we will fail the first fundamental requirement, that $v_i > v_{i+k}$ for all $i, k \in \mathbb Z^+$.

The mean GP by pick number and mean PS by pick number metrics are two of the metrics we will use in the Model chapter.
